Transfer of Learned Opponent Models in Zero Sum Games

### 2021-06-04 Report/draft_upgrade_v1.Rmd
added report in papaja and softmax to BCH models

abstract: |
  Enter abstract here. Each new line herein must be indented, like this line.

  Human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and 
  causal models of the environment. One way to build abstract representations of the environment when the task involves interactions with others is to build 
  a model of the opponent that may inform what actions they are likely to take next. In this study, we propose to explore opponent modelling and its transfer with the use of
  computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. Through two differentiated experiments, we find that participants can
  deviate from Nash equilibrium play and learn to adapt to the opponent strategy and exploit it. Moreover, we showed that participants transfer their learning to new games and that
  the transfer is moderated by the level of  sophistication of the opponent. Computational modelling shows that it is likely that players start each game using a model-based
  learning strategy that facilitates generalisation and opponent model transfer, but then switch to behaviour that is consistent with a  model-free learning strategy in the latter
  stages of the interaction.

@@

$$ A_{i}^{j}(t) =  \frac{\phi . N(t-1) . A_{i}^{j}(t-1) + [ \delta + (1-\delta) . I(s_{i}^{j},s_{i}(t))] . \pi_{i}(s_{i}^{j},s_{-i}(t)) } {\phi . N(t-1) + 1} $$

Here, I(x,y) is the indicator function equal to 1 if $x = y$ and 0 otherwise. A simple way to think about this update rule is that attractions are multiplied by a parameter that represents experience ($N(t-1)$) which is itself decaying by a weight $\phi$. The result is added to either the pay-off received (when the indicator function is 0), or to $\delta$ times the foregone pay-off (when indicator function is 1).   We can see that setting $\delta = 0$ leads to reinforcement of past actions, while positive and high delta parameters make the update rule take into account foregone pay-offs, which is similar to weighted fictitious play [@cheung1994learning]. 
While the assumption in expanding the update rule above is that $\phi$ and $\delta$ are free parameters [@camerer1997experience], the self-tuning aspect of the model comes from the fact that these are now self-tuned using the formulas expanded in [@ho2004economics].

### ToM models

In this set of models, we assume that participants have a belief that the opponent is a level-k agent, with uniform probability of the level k, and use evidence of past play to update their beliefs in a Bayesian way about the true value of k. We use values of k  in ${0,1,2}$. These models also assume the opponent can deviate from these level-k strategies and play randomly with probability $\theta$, a parameter to be fit. We distinguish between multiple ToM models based on their ability to keep what was learned about the opponent in memory and hence facilitate transfer. In a No-Between-Transfer (NBT) model, players have no memory of what was learned about the opponent and start every new game assuming each level-k has equal probability. In the context of Experiment 2 where players face both opponents, this model assumes that participants transfer learning within the same game, from the first to the second interaction with the opponent, but are not able to transfer that learning to new games (So within but no between transfer)

Conversely, In a Between-Transfer model (BT), players are assumed to keep in memory what was learned about the type of opponent faced (vector of probabilities of level-k) and use that at the beginning of each new game. In the context of experiment 2, we still assume that if between transfer is present, then within game transfer is also present (from first to second interaction).

We fit the above two models to experiment 1 data. In experiment two, on top of these two models, we fit another model in which all stages of the game and all new games start with a fresh uniform probability of level-k opponent (NT), so no within or between opponent model learning transfer.

### Estimation and model comparison
In both experiments, all models were fit to each participant data, with optimal parameters being estimated using maximum likelihood. Using information criteria based Bayesian model comparison (BIC), the best fitting model for each participant was chosen and we compared the number of participants whose behavior was best explained by each model.

@@

Figure\ \@ref(fig:exp1-comp-models)  shows the results for experiment 1: we can see that while some participant's learning behavior was either random or explained by some of the base models, a significant number of participants in experiment one had learning most consistent with Q-learning with states defined by last round play.

@@

Table \ \@ref(tab:exp1weightstable) shows the model BIC weights  as well as the number of participants best fit by each model.

@@

```{r, include=TRUE}

@@

ggplot(data_long_post1,aes(x=round,y=probability,colour=strategy)) +
  geom_line() +
  facet_grid(game~condition)

@@

In experiment 1, the HMM model with switching fits significantly better than the non-switching one ($p < .001$). This is further statistical evidence in favour of the hypothesis that participants switch between strategies. In order to understand at which stage of the games the switching might happen, and whether there are any differences between games and type of opponents faced, we plot in Figure\ \@ref(fig:exp1-posteriors-plot) the average (across participants) posterior probabilities of each state (strategy), as a function of trial and opponent faced. The posterior probability is the probability that an observation comes from a component distribution a posteriori, i.e. given the value of the observation. In the first experiment, we can see from the plots of fwg and numbers games for level-1 opponent that although the likelihoods are very close, the posterior probability of the Bayesian model with transfer is slightly higher than that of the QLS model in the very early rounds, but decreases rapidly while the posterior probability of the QL-learning with states models keeps increasing.

@@

The results of our first online experiment show that the majority of participants learn to adapt to the opponent strategy over multiple interactions and generalise this learning to the similar game. We used results on very early rounds for measuring transfer as they are unlikely to be tainted by any within game learning. Using this approach, we showed that transfer to the more dissimilar game was modulated by the degree of sophistication of the agent, with evidence for transfer when players face the less sophisticated agent but not the more sophisticated one.

@@

We will proceed by considering multiple potential answers to this question. First, maybe players simply learn spatial heuristics that allow them to perform better than chance. An example is a spatial heuristic that consists of choosing "weapons" in a particular order. For instance, it is possible to keep winning against a level-k opponent by choosing actions in a particular spatial order such as cycling through them from left to right. This was one of the weaknesses in the design of experiment one, as it was indeed possible using very simple spatial sequences to beat the opponent on most rounds. We took this into account in designing experiment two by randomly shuffling the spatial order of action choices in each round. Still, the learning and conclusions were similar, so this could not explain both learning and transfer in experiments one and two.

A second possible hypothesis for learning the opponent’s strategy is the use of simple rules based on last round play (for instance, I play scissors whenever opponent played rock in last round, or whenever the last round play was rock/scissors, I should play paper in this round, etc...). Our Q-learning with states as prior-round play model is a good proxy for this type of strategies. While this approach certainly seemed to be the best fit for some player's behavior, it is unsatisfactory in explaining some of the learning transfer evidence we showed. Indeed, learning the best action in a particular state is not transferable to a new game since the state space is different and there is no single mapping between the state spaces of the initial and latter games. These rules would therefore need to be learned anew in the latter game which is inconsistent with above chance performance in very early rounds.

@@

Our computational modelling allowed us to delve deeper into what might be driving the observed learning transfer. Initial modelling of observations using all available data seems to indicate that the most likely model was a Q-learning type model. However, as we argued above, that would be inconsistent with the evidence for learning transfer. Breaking down likelihoods by trial and fitting a hidden markov model to the data with states being the various strategies that participants are assumed to be using, we showed evidence for within game switching of strategies. Participants start the early rounds of a new game acting in a way consistent with a Bayesian Theory of mind level, which would be accurate and generalisable but computationally expensive. However, as trials continue, participants seem to switch to a habitual type of learning (QL-models).

>>>>

abstract: |
  Human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and 
  causal models of the environment. One way to build abstract representations of the environment when the task involves interactions with others is to build 
  a model of the opponent that may inform what actions they are likely to take next. In this study, we propose to explore opponent modelling and its transfer with the use of
  computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. Through two differentiated experiments, we find that participants can
  deviate from Nash equilibrium play and learn to adapt to the opponent strategy and exploit it. Moreover, we showed that participants transfer their learning to new games and that
  the transfer is moderated by the level of  sophistication of the opponent. Computational modelling shows that it is likely that players start each game using a model-based
  learning strategy that facilitates generalisation and opponent model transfer, but then switch to behaviour that is consistent with a  model-free learning strategy in the latter
  stages of the interaction.

@@

$$ A_{i}^{j}(t) =  \frac{\phi \times N(t-1) \times A_{i}^{j}(t-1) + [ \delta + (1-\delta) \times I(s_{i}^{j},s_{i}(t))] \times \pi_{i}(s_{i}^{j},s_{-i}(t)) } {\phi \times N(t-1) + 1} $$

Here, $I(x,y)$ is the indicator function equal to 1 if $x = y$ and 0 otherwise. A simple way to think about this update rule is that attractions are multiplied by a parameter that represents experience ($N(t-1)$) which is itself decaying by a weight $\phi$. The result is added to either the pay-off received (when the indicator function is 0), or to $\delta$ times the foregone pay-off (when indicator function is 1).   We can see that setting $\delta = 0$ leads to reinforcement of past actions, while positive and high delta parameters make the update rule take into account foregone pay-offs, which is similar to weighted fictitious play [@cheung1994learning]. While the assumption in expanding the update rule above is that $\phi$ and $\delta$ are free parameters [@camerer1997experience], the self-tuning aspect of the model comes from the fact that these are now self-tuned using the formulas expanded in [@ho2004economics].

## ToM models

In this set of models, we assume that participants start with a prior belief that the opponent is a level-k agent, with uniform probability of the level $k \in \{ 0, 1, 2\}$, and use evidence of past play to update their beliefs in a Bayesian way about the true value of k. More formally, if we define $D_{t}$ as data available at time $t$ then the probability of the opponent being a level-k is updated using:

$$ P(level=k | D_{t})  \propto  P(D_{t}|level=k ) \times P(level=k)$$

These models also assume the opponent can deviate from these level-k strategies and play randomly with probability $\theta$, a parameter to be fit. We also assume that participants base their actions on their beliefs about the level of the opponent, so actions get chosen probabilistically using a softamx choice rule with probabilities of each opponent level as inputs. We distinguish between multiple ToM models based on their ability to keep what was learned about the opponent in memory and hence facilitate transfer. In a No-Between-Transfer (NBT) model, players have no memory of what was learned about the opponent and start every new game assuming each level-k has equal probability. In the context of Experiment 2 where players face both opponents, this model assumes that participants transfer learning within the same game, from the first to the second interaction with the opponent, but are not able to transfer that learning to new games (within but no between transfer).Conversely, In a Between-Transfer model (BT), players are assumed to keep in memory what was learned about the type of opponent faced (vector of probabilities of level-k) and use that at the beginning of each new game. In the context of experiment 2, we still assume that if between transfer is present, then within game transfer is also present (from first to second interaction).  In experiment two, on top of these two models, we fit another model in which all stages of the game and all new games start with a fresh uniform probability of level-k opponent (NT), so no within or between opponent model learning transfer.

## Estimation and model comparison
In both experiments, all models were fit to each participant data, with optimal parameters being estimated using maximum likelihood. Using information criteria based Bayesian model comparison (BIC), the best fitting model for each participant was chosen and we compared the number of participants whose behavior was best explained by each model.

In experiment 1, we fit a total of 5 models: A baseline model assuming random play (Nash), a Bayesian Theory of mind model allowing transfer betwen games (TOM_BT) and another with no transfer between games (TOM_NBT), as well as a Q-learning model with state space consisting of last round play (QL_states), and finally a self tuning EWA model with the same state space (STEWA_states). In experiment 2, because participants were interacting with each opponent twice within each game, we allowed models to transfer learning within games between the first and second interactions. Therefore on top of the two ToM models mentionned above (ToM_BT and Tom_NBT), we also allow the Q-learning model with states to either have the ability to transfer learning between the first and second interaction (QLS_Tr) or not (QLS_NT). Likewise, we fit both a self tuning EWA model with transfer between stages of the same game (STEWA_Tr) or without transfer (STEWA_NT). Counting the base model wiht random play (Nash) we therefore fit a total of 8 models in experiment 2.

@@

Figure\ \@ref(fig:exp1-comp-models)  shows the results for experiment 1: we can see that while some participant's learning behavior was either random or explained by some of the base models, a significant number of participants in experiment one had learning most consistent with Q-learning with states defined by last round play. 
Table \ \@ref(tab:exp1weightstable) shows the model BIC weights  as well as the number of participants best fit by each model.

@@

@@

@@

p <- ggplot(data_long_post1,aes(x=round,y=probability,colour=strategy)) +
  geom_line() +
  facet_grid(game~condition)

p

# p + theme(legend.position = c(0.3,0.5)
#           ,legend.background = element_rect(fill = "#ffffffaa", colour = NA))

@@

In experiment 1, the HMM model with switching fits significantly better than the non-switching one ($p < .001$). This is further statistical evidence in favour of the hypothesis that participants switch between strategies. In order to understand at which stage of the games the switching might happen, and whether there are any differences between games and type of opponents faced, we plot in Figure\ \@ref(fig:exp1-posteriors-plot) the average (across participants) posterior probabilities of each state (strategy), as a function of trial and opponent faced. The posterior probability is the probability that an observation comes from a component distribution a posteriori, i.e. given the value of the observation. In the first experiment, we can see from the plots of FWG and numbers games for level-1 opponent that although the likelihoods are very close, the posterior probability of the Bayesian model with transfer is slightly higher than that of the QLS model in the very early rounds, but decreases rapidly while the posterior probability of the QL-learning with states models keeps increasing.

@@

The results of our first experiment show that the majority of participants learn to adapt to the opponent strategy over multiple interactions and generalise this learning to the similar game. We used results on very early rounds for measuring transfer as they are unlikely to be tainted by any within game learning. Using this approach, we showed that transfer to the more dissimilar game was modulated by the degree of sophistication of the agent, with evidence for transfer when players face the less sophisticated agent but not the more sophisticated one.

@@

We will proceed by considering multiple potential answers to this question. A possible hypothesis for learning the opponent’s strategy is the use of simple rules based on last round play (for instance, I play scissors whenever opponent played rock in last round, or whenever the last round play was rock/scissors, I should play paper in this round, etc...). Our Q-learning with states as prior-round play model is a good proxy for this type of strategies. While this approach certainly seemed to be the best fit for many player's behavior, it is unsatisfactory in explaining some of the learning transfer evidence we showed. Indeed, learning the best action in a particular state is not transferable to a new game since the state space is different and there is no single mapping between the state spaces of the initial and latter games. These rules would therefore need to be learned anew in the latter game which is inconsistent with above chance performance in very early rounds.

@@

Our computational modelling allowed us to delve deeper into what might be driving the observed learning transfer. Initial modelling of observations using all available data seems to indicate that the most likely model was a Q-learning type model. However, as we argued above, that would be inconsistent with the evidence for learning transfer. Breaking down likelihoods by trial and fitting a hidden Markov model to the data with states being the various strategies that participants are assumed to be using, we showed evidence for within game switching of strategies. Participants start the early rounds of a new game acting in a way consistent with a Bayesian Theory of mind level, which would be accurate and generalisable but computationally expensive. However, as trials continue, participants seem to switch to a habitual type of learning (QL-models).
